---
title: "Data Analysis Assignment"
author: "Elijah Verdoorn"
date: "September 25, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(class)
```

# Classification
## Setup

```{r}
maxKVal <- 100
means <- matrix(nrow = maxKVal, ncol = 1)
```
Build a matrix in which to store the means that get calculated, as well as the limiter of the `for` loop.

## Main Loop

```{r}
for (i in 1:maxKVal) {
  kval <- i # number of neighbors
  
  train.df <- read.csv("adult.csv") # read in the training data
  test.df <- read.csv("adult_test.csv") # read in the testing data
  
  response.predict <- knn(train.df[,c("Age", "race", "sex")], 
                          test.df[,c("Age", "race", "sex")], 
                          train.df[,c("X..50k")], k = kval) # build the KNN model
  response.predict # print the testing results
  
  test.df <- mutate(test.df, class.pred = response.predict)
  # how did we do? calculate the error rate
  means[i] <- mean(test.df[,c("X1")] != test.df["class.pred"])
}
```
The main loop, which builds `maxKVal` models and calculates the error for those models.

## Analysis

```{r}
min(means) # at k = 78
```
Find the minimum error, in this case that occurs around `k = 40`.

# Regression
## Setup
```{r}
# get the dataset
data <- read.csv("winequality-white.csv")
degrees <- 20 # the number of degrees to test
results.df <- data.frame("degree" = integer(), "train" = double(), "test" = double())
```

## Main Loop
```{r}
for(i in 1:degrees) {
  n <- nrow(data) # get the number of rows
  train <- sample(1:n, n/2, rep = F) # get subset of the data for training
  test <- setdiff(1:n, train) # and one for testing
  
  train.df <- data[train, c("pH", "alcohol")] # set up the data frame for training
  test.df <- data[test, c("pH", "alcohol")] # set up the data frame for testing
  
  mod <- lm(pH~poly(alcohol, i), data = train.df) # build the linear model with degree i
  pred.train <- predict(mod, newdata = train.df) # traing the model
  train.df <- mutate(train.df, pred = pred.train) # add the training data to the data frame
  
  pred.test <- predict(mod, newdata = test.df) # test the model
  test.df <- mutate(test.df, pred = pred.test) # add the test data to the data frame
  
  mse.train <- with(train.df, mean((pH - pred)^2)) # calculate information about the model
  mse.test <- with(test.df, mean((pH - pred)^2))
  c(mse.train, mse.test) # print the information about the model to the console
  results.df <- rbind(results.df, c(i, mse.train, mse.test))
}
```
Calculate the models, then add information about that degree to the data frame.

## Results
```{r}
head(results.df)
```