---
title: "SPAM and Trees"
author: "Matt Richey"
date: "October 12, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r,message = F}
library(tidyverse)
library(tree)
```

Predict SPAM values. Need train/test data. We will use
cross-validated tree with pruning



Read in the  SPAM data set (from UCI)
```{r}
spam.df <- read.csv("SPAM.csv",header=T)
head(spam.df)

```

Make sure response is a Logical

```{r}
spam.df <- mutate(spam.df,IsSpam=factor(ifelse(IsSpam=="TRUE",1,0)))
with(spam.df,table(IsSpam))
str(spam.df$IsSpam)

```

Build  train and test data

```{r}
nn <- nrow(spam.df)
train <- sample(1:nn,nn/2,rep=F)
spamTrain.df <- spam.df[train,]
spamTest.df <- spam.df[-train,]


```

Our first tree...make is sorta deep since we're going to prune it
later

```{r}
spam.tree <- tree(IsSpam~.,data=spamTrain.df,
                  control=tree.control(nrow(spamTrain.df),mindev=0.001))
plot(spam.tree)
text(spam.tree,cex=0.5)


```

Cross validate to find the optimal size

```{r}
spam.cv <- cv.tree(spam.tree,FUN=prune.misclass)
plot(spam.cv)
spam.cv

```

Dig out the optimal values..need to be careful here to make sure
getting the optimal value
I like to reverse the values to make sure we get smallest

```{r}
sz <- rev(spam.cv$size)
(best.sz <- sz[which.min(rev(spam.cv$dev))])


```

Prune the tree and plot it.

```{r}
spam.prune=prune.tree(spam.tree,best=best.sz)
spam.prune
plot(spam.prune)
text(spam.prune,cex=0.5)

```


Now we can make predictions on the Test data
```{r}
preds.test <- predict(spam.prune,newdata=spamTest.df,type="class")
table(preds.test)

```

This will give you the probability of each class..Use this for an
ROC curve or threshholding in general

```{r}
probs.test <- predict(spam.prune,newdata=spamTest.df,type="vector")
hist(probs.test[,1],breaks=25)

```

How well did we do?

```{r}
with(spamTest.df,table(preds.test,IsSpam))
with(spamTest.df,mean(preds.test!=IsSpam))

```

Not bad.  Now predict on the Training Data.. better

```{r}
preds.train <- predict(spam.prune,newdata=spamTrain.df,type="class")
with(spamTrain.df,table(preds.train,IsSpam))
with(spamTrain.df,mean(preds.train!=IsSpam))


```

## Bootstrap the data


```{r}
m.train <- nrow(spamTrain.df)
m.test <- nrow(spamTest.df)
B <- 40
bootVals <- matrix(nrow=m.test,ncol=B)

preds.oob.mat <- matrix(nrow = m.train, ncol = B)

for(b in 1:B){
  print(b)
    boots <- sample(m.train,m.train,rep=T)
    spamBoot.df <- spamTrain.df[boots,]
    spam.tree <- tree(IsSpam~.,data=spamBoot.df,
                      control=tree.control(nrow(spamBoot.df),
                                           mindev=0.005))
    
    spam.cv <- cv.tree(spam.tree)
    sz <- rev(spam.cv$size)
    bs <- sz[which.min(rev(spam.cv$dev))]
    bs<-15
    spam.prune=prune.misclass(spam.tree,best=bs)
    ###
    preds.test <- predict(spam.prune,
                           newdata=spamTest.df,
                           type="class")
    bootVals[,b] <-as.numeric(as.character(preds.test))
    
    # For the Assignment listed below
    oob.vals <- setdiff(1:m.train, boots) # get the OOB values
    spamOOB.df <- spamTrain.df[oob.vals,]
    preds.oob <- predict(spam.prune, newdata = spamOOB.df, type = "class")
    preds.oob.mat[oob.vals,b] <- as.numeric(as.character(preds.oob))
}


dim(bootVals)
```


Now we can see how well boot strapping did.
First we take the mean of hte row, then compare with 1/2. 
```{r}
vals <- apply(bootVals,1,mean)
preds.boot <- as.numeric(vals>0.5)
with(spamTest.df,table(IsSpam,preds.boot))
with(spamTest.df,mean(IsSpam != preds.boot))

```


This can be done (better) with the built in library/function randomForest.


```{r}
library(randomForest)
p <- ncol(spamTrain.df)-1

spam.bag <- randomForest(IsSpam~.,
                         data=spamTrain.df,
                         #mtry=sqrt(p), ## For Random Forest
                         mtry=p, ##For BAG
                         ntree=100)
preds.bag <- predict(spam.bag,newdata=spamTest.df)
with(spamTest.df,table(IsSpam,preds.bag))
with(spamTest.df,mean(IsSpam!=preds.bag))
```


This shows the OOB error rate, the False Positive, and False Negative Error rates as a function of the number of trees.
```{r}
plot(spam.bag)
```

We can see how the random forest works by looking at its output.
Among other things, we can see the OOB confusion matrix and error rates.

```{r}
spam.bag
```

Look at variable importance, that is, how much of does each variable contribute to the decrease in the Error Rate (Gini Factor) at each split.


Turn importance into a data frame for plotting.
```{r}
vv<-spam.bag$importance
(varImp.df<-data.frame(var=as.character(row.names(vv)),GiniDec=vv[,1]))
```


Put the variable variable in order of importance.

```{r}
varOrder<-varImp.df %>% 
  arrange(GiniDec) %>% 
  with(var)

```

Fix the factors in the order of importance.

```{r}
varImp.df<-varImp.df %>% 
  mutate(var=factor(var,levels=varOrder))

```


Plot it....

```{r}
ggplot(varImp.df[1:20,])+
  geom_bar(aes(x=var,GiniDec),stat="identity",fill="red",color="black",
  width=0.5)+
  scale_y_continuous("Gini Decrease")+
  coord_flip()+
  ggtitle("Top Twenty Variables in SPAM")

```



## Computing Out of Bag error

The out of bag (OOB) error is derived by using the values that are not in the boot strap as the training data. 

* Goal: Using our "by hand" bootstrap, compute the OOB error rate on the spamTrain data. How close does this come to predicting the the test error rate computed on spamTest?

* Extra: Also compute the OOB False Positive and False Negative rates. This should be easy.

* Extra Extra: Plot these rates as a function of hte number of trees used int the bootstrap. Doing so will recreate the plot of the random forest.

*Plan*: As we boot strap, keep track of the OOB values as the test data. After creating the boot strap tree, predict on the OOB values. Store the results in a matrix

