---
title: "Modeling Concrete Strength"
author: "Matt Richey"
date: "September 29, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction
We look at modeling a data set for concrete strength using both linear regression and KNN.

Load the libraries, including FNN for the knn.reg.
```{r, message=F}
library(ggplot2)
library(dplyr)
library(tidyr)
library(FNN)
```

Read the data
```{r}
concrete.df<-read.csv("concrete.csv")
```
```{r}
names(concrete.df)
```
SLUMP, FLOW, and Strength are all potential response variables, only keep Strength.

No is just a row number.
```{r}
concrete.df <- concrete.df %>% 
  select(-SLUMP,-FLOW,-No)

```


Take a look at remaining combinations.
```{r}
pairs(concrete.df)
```

There appears to be a potential for a relationship between Cement, Slag and Strength.

Quick Plots


```{r}
with(concrete.df,plot(Cement,Strength))
```

```{r}
with(concrete.df,plot(Slag,Strength))
```

Look a little closer at Slag. There are a lot of zero values.
```{r}
with(concrete.df,hist(Slag,breaks=25))
```

Create a qualitative predictor from Slag
```{r}
concrete.df<-mutate(concrete.df,LowSlag=Slag==0)
```



## Model Strength with Concrete and LowSlag/Slag
 Look at the combination of the quantitative and 
qualitative predictors
```{r}
ggplot(concrete.df,aes(Cement,Strength,color=LowSlag))+
  geom_point()

```


A basic linear model with two predictors

```{r}
mod1<-lm(Strength~Cement+LowSlag,data=concrete.df)
```



The model summary

```{r}
summary(mod1)
```
How do we read this??

* Increasing Cement increases the Strength
* LowSlag==TRUE also appears to increase the strength
* Both these predictors are statistically significant

Now make predictions on the training set

```{r}
pred<-predict(mod1)
concrete.df$pred<-pred

```

The visual inspection of predictions

```{r}
ggplot(concrete.df,aes(Cement,Strength,color=LowSlag))+
  geom_point()+
  geom_line(aes(Cement,pred))+
  ggtitle("Concrete: Linear Model")

```


Now add an interaction between Concrete and LowSlag
```{r}
mod2<-lm(Strength~Cement+LowSlag+Cement:LowSlag,data=concrete.df)
summary(mod2)
```

The Cement:LowSlagTRUE coefficient is positive (but not significant). This is saying there is very weak evidence that in the presence of LowSlag, that the Cement effect is even stronger.



Plot this

```{r}
concrete.df$pred<-predict(mod2)
ggplot(concrete.df,aes(Cement,Strength,color=LowSlag))+
  geom_point()+
  geom_line(aes(Cement,pred))  

```

Using geom_smooth(method=lm) gives the same graph.
```{r}
ggplot(concrete.df,aes(Cement,Strength,color=LowSlag))+
  geom_point()+
  geom_smooth(method=lm,se=F)

```


# Prediction

Let's focus on just Cement
```{r}
mod0<-lm(Strength~Cement,data=concrete.df)
concrete.df$pred<-predict(mod0)
summary(mod0)



```

The plot
```{r}
ggplot(concrete.df,aes(Cement,Strength))+
  geom_point()+
  geom_line(aes(Cement,pred))  

```


Let's do some a "Feature expansion" : add quadratic term in Cement.


```{r}
mod2<-lm(Strength~Cement+I(Cement^2),data=concrete.df)
concrete.df$pred<-predict(mod2)

```


And the plot...
```{r}
ggplot(concrete.df,aes(Cement,Strength))+
  geom_point()+
  geom_line(aes(Cement,pred))  
```


This is same adding a new predictor equal to Cement^2
```{r}
concrete.df<-concrete.df %>% 
  mutate(Cement2=Cement^2)
mod2x<-lm(Strength~Cement+Cement2,data=concrete.df)

```

The summaries are evidence that they are the same.

```{r}
summary(mod2x)
summary(mod2)

```




More feature expansion: cubic term

```{r}
mod3<-lm(Strength~Cement+I(Cement^2)+I(Cement^3),data=concrete.df)
summary(mod3)
concrete.df$pred<-predict(mod3)
```
The plot of the cubic fit.
```{r}
ggplot(concrete.df,aes(Cement,Strength))+
  geom_point()+
  geom_line(aes(Cement,pred))  
```


More feature expansion: cubic term: quartic term
```{r}
mod4<-lm(Strength~Cement+I(Cement^2)+I(Cement^3)+I(Cement^4),
         data=concrete.df)
summary(mod4)
```



Notice that none of the terms are significant now

```{r}
concrete.df$pred<-predict(mod4)
ggplot(concrete.df,aes(Cement,Strength))+
  geom_point()+
  geom_line(aes(Cement,pred))  
```





More feature expansion: Add *any number of powers of
Cement

To do so, 
Create formula string of the form
  "Strength~Cement+I(Cement^2)+I(Cement^3)+I(Cement^4)""
Here's how to do it.

Fix the degree
```{r}
maxDeg<-4
```
Begin with a starter string and iteratively add more terms.
```{r}
formStr<-"Strength~Cement" #starter string
for(d in 2:maxDeg){
  formStr<-sprintf("%s+I(Cement^%s)",formStr,d)
}
formStr
```



Convert string to formula using function
**formula**
```{r}
modMax<-lm(formula(formStr),
         data=concrete.df)
summary(modMax)

```

Now we can easily  model using any degree we want.

```{r}
concrete.df$pred.lm<-predict(modMax)
ggplot(concrete.df,aes(Cement,Strength))+
  geom_point()+
  geom_line(aes(Cement,pred.lm))  

```



Natural question:
 * Which degree is best? Use a train/test  approach to find optimal value. 



## Nearest Neighbors:  KNN for regression

Use library(FNN)
and check out
?knn.reg


For knn.reg, need to use matrices again. 

Here we go bulding matrices of the proper dimensions from a data frame

The data
```{r}
cement<-with(concrete.df,Cement)
resp<-with(concrete.df,Strength)

```
The dimensions
```{r}
dim(cement)<-c(nrow(concrete.df),1)
dim(resp)<-c(nrow(concrete.df),1)

```

Set the number of neighbors
```{r}
kval<-10
```


Use original data as test data, for comparison to  linear regression approach
```{r}
mod.knn<-knn.reg(cement,NULL,resp,k=kval)
```
Here are the predictions
```{r}
concrete.df$pred.knn<-mod.knn$pred
```


And a plot...
```{r}
ggplot(data=concrete.df)+
  geom_point(aes(Cement,Strength))+
  geom_step(aes(Cement,pred.knn),color="red")+
  geom_line(aes(Cement,pred.lm),color="blue")+
  ggtitle(sprintf("KNN with k = %s, Linear model with degree=%s",kval,maxDeg))

```

Build KNN.reg on an arbitrary test data set. Here I'll use equally spaced values over the range of the Cement values.
```{r}
nn<-75
rr<-with(concrete.df,range(Cement))
test<-seq(rr[1],rr[2],len=nn)
dim(test)<-c(nn,1)
```

Now use knn.reg with the test data set.
```{r}
mod.knn<-knn.reg(cement,test,resp,k=kval)
```

Here's a way to combine the train and test data in a single plot.
```{r}
ggplot()+
  geom_point(data=concrete.df,aes(Cement,Strength))+
  geom_step(data=NULL,aes(x=test,y=mod.knn$pred),color="red")
```

# Assignment
 Comparison of Linear Models and KNN as a predictor of Concrete 
 Strength.
 
* Stage 1: Single predictor: Cement --> Strength
  Which one is better...knn (kval=?) or lm (maxDeg=?)
* Stage 2: Two predictors: Cement+Slag (continuous) --> Strength
 Same question
* Stage 3: Add an additional quantitative predictor. Does this improve  the prediction power of either method? 


For your analysis, use multiple random train+test combinations to
 evaluate your MSE. Each time, use half the data used to train, the other half used to test.


